---
title: "Biophybreak Wrapper Function Usage"
author: "Erik Lundgren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biophybreak Wrapper Function Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  out.width = "100%",
  comment = "#>"
)
```

## Introduction

This vignette is intended to provide an explanation of the biophybreak workflow with the wrapper functions when using real data.
Real data is not actually available for this example, but I will do some transformations to make it more similar in format to real data
and demonstrate the functions that you would typically use with real data.

## load packages and set pRNG seed

```{r packages}
library(biophybreak)
library(parallel)

seed <- 1234
set.seed(seed)
```

## Simulate several transmission histories

```{r simulate_transmission_histories}
#number of individuals in each cluster
nInds <- c(2, 5, 8)

#number of transmission clusters
nClust <- length(nInds)

#time betweeen infection and sampling is gamma distributed with mean 2 and sd 1.5
infection.ages <- lapply(nInds, FUN = rgamma, shape = 16/9, rate = 8/9)

#type of distribution for transmissibility profile
#(relative chance of transmission based on how long an individual has been infected)
trans.prof.type <- "function"
transmissibility.profile <- stepfun(x = c(0, 0.5), y = c(0, 3, 1)) #three times more likely to transmit in first 6 months

#declare list for transmission histories
tt <- vector(mode = "list", length = nClust)
#generate transmission histories
for(i in seq_len(nClust)){
  tt[[i]] <- rtrans_tree(nInd = nInds[i], 
                         infection.ages = infection.ages[[i]],
                         #trans.prof.type = trans.prof.type, 
                         #transmissibility.profile = transmissibility.profile
                         gen.dist.type = trans.prof.type, 
                         gen.dist = transmissibility.profile)
}

#declare list for simulated data
sim_data <- vector(mode = "list", length = nClust)
#simulate phylogenies and sequences
for(i in seq_along(nInds)){
  sim_data[[i]] <- sim.coal.phybreak(tt[[i]], 
                                     a = rep(5, nInds[i]),
                                     b = rep(5, nInds[i]),
                                     mut_model = "pol")
}
```

## Plot simulated transmission histories

```{r plot_sim_histories, fig.width=6, fig.height=4}
plotPhyloTrans(sim_data[[1]])
plotPhyloTrans(sim_data[[2]])
plotPhyloTrans(sim_data[[3]])
```

## Simulate biomarker values

```{r}
#collapse infection ages into single vector
infection.ages.vector <- do.call("c", infection.ages)
#draw random biomarkers based on infection ages
biomarkers <- sim.biomarkers(infection.ages.vector)
```

## Convert simulated data to a format more similar to real data

```{r}
#put all sampling times (relative to first infection time in each cluster) in a vector
all_t_sam <- unlist(lapply(tt, FUN = function(tt) tt$t_sam))

#find approximate dates for sampling times assuming the earliest infected individual in the cluster is infected 2000-01-01
t_sam_dates <- as.Date(all_t_sam*365.25, origin = "2000-01-01")

#add cluster number to patient names
#paste(1, tt[[1]]$names, sep = "_")
patient_ID <- unlist(lapply(tt, FUN = function(tt) tt$names))

#vector for the cluster name for each individual
#repeat the cluster ID a number of times equal to the number of individuals in the cluster
cluster_ID <- unlist(mapply(seq_along(nInds), FUN = rep, nInds))

#extract simulated sequences from transmission history simulation data
extract_seqs <- function(sim_data, tt){
  nInd <- length(tt$t_sam)
  seqs <- vector(mode = "list", length = nInd)
  for(i in seq_len(nInd)){
    seqs[[i]] <- sim_data$sequences[i,,drop = FALSE]
  }
  #put back into original order instead of order based on sampling time
  seqs <- seqs[order(order(tt$t_sam))]
  return(seqs)
}
seqs <- mapply(FUN = extract_seqs, sim_data, tt)
#change to single list instead of list of lists
seqs <- do.call("c", seqs)

#sequence names
seq_names <- lapply(seqs, FUN = names)
```

## Put data into dataframe

The `prepare.HIV.data` function combines all of the biomarker, sequence, and demographic data 
needed for later inference into a data frame.
Here we also use `find_infection_age_distributions = TRUE` in order to run the multiple biomarker model
and calculate the distributions for the amount of time between infection and sampling.
It is also possible to define your own columns in the data frame by passing new parameters to the function.

```{r prepare.HIV.data}
df <- prepare.HIV.data(patient_ID = patient_ID,
                       last_neg_test_date = NULL,
                       first_pos_test_date = t_sam_dates, 
                       ART_start_date = NULL,
                       BED_dates = t_sam_dates,
                       BED = biomarkers[,"BED"],
                       LAg_dates = t_sam_dates,
                       LAg = biomarkers[,"LAg"],
                       CD4_dates = t_sam_dates,
                       CD4 = biomarkers[,"CD4"],
                       seq_dates = t_sam_dates,
                       seqs = seqs,
                       seq_names = seq_names,
                       pol_override = biomarkers[,"pol"], #you would normally not use this
                       pol2_dates = t_sam_dates,
                       pol2 = biomarkers[,"pol2"], 
                       VL_dates = NULL,
                       VL = NULL,
                       ART_cutoff_delay = 3, #in days
                       cluster_ID = cluster_ID,
                       gender = NULL,
                       age_at_sampling = NULL,
                       birth_location = NULL,
                       suspected_infection_location = NULL,
                       risk_group = NULL,
                       aids_diagnosis_date = NULL,
                       death_date = NULL,
                       date_format = "%Y-%m-%d",
                       find_infection_age_distributions = TRUE,
                       prior.type = 1,
                       user.prior.pdf = list(x = c(0, 10), y = c(1/10, 1/10)),
                       n.adapt = 1e4, 
                       n.burn = 1e4, 
                       n.iter = 1e4,
                       seed = sample(2^31-1, 1))
```

## Specify additional parameters for inference
Some additional parameters must be specified in order to run biophybreak, 
such as the within-host model of effective population size growth, 
transmissibility profile (how the likelihood of transmitting to a new individual changes over time since infection),
and the mutation modelused for sequence likelihood.

```{r}
inputs <- prepare.biophybreak.data(df, 
                                   wh.model = "linear", #type of function for growth of within-host effective pop size
                                   wh.bottleneck = "wide", #bottleneck size of transmission
                                   prior.wh.level.mean = 0.1, #prior for bottleneck size
                                   gen.type = "function", #type of prior for transmissibility profile
                                   gen.density = stepfun(c(0, 0.5), c(0, 3, 1)), #transmissibility profile
                                   post.sam.trans.rate = 1, #factor indicating how much less likely transmission is after sampling
                                   est.gen.mean = FALSE, #set to false if not using a gamma distribution for gen.type
                                   use.pml = TRUE, #whether to use the phangorn::pml function for sequence likelihood 
                                   mut.model = "pol", #the mutation model to use with phangorn::pml
                                   run.names = "automatic",
                                   overall.seed = sample(2^31-1,1)) #pRNG seed
```

## Run biophybreak

The run.biophybreak function is a wrapper for 
the `phybreak`, `burnin_phybreak`, and `sample_phybreak` functions. 
If you need to run transmission history inference on multiple  transmission clusters, 
you will most likely want to parallelize the computation with mclapply or parLapply.
This will take MCMC samples until either the desired effective sample size for all parameters
or the maximum number of iterations is reached.

```{r}
cl <- makePSOCKcluster(min(nClust, detectCores()))
setDefaultCluster(cl)
clusterEvalQ(NULL, library(biophybreak))
MCMCs <- parLapply(NULL,
                   inputs, 
                   fun = run.biophybreak,
                   burn_iter = 1e4, 
                   iter_max = 2e5, 
                   iter_block = 1e4, 
                   thin = 1, 
                   ESS_target = 200,
                   save = FALSE,
                   outdir = NULL)
stopCluster(cl)
```

## Plot posterior probabilities and consensus trees

The `phybreak.plot.triple` function is a good way to get an overview of the inference results for each cluster.
It plots the maximum parent credibility transmission history, 
the inferred infection times from the multiple biomarker model, 
and the posterior probabilities of the potential sources for each individual.
If the true history is known, such as when doing a simulation study, 
the true history can also be plotted, and the posterior probability plot shows the true source with a darker outline.

```{r plot_inference, fig.width = 7.5, fig.height= 10}
#assuming true history is unknown
for(i in seq_along(MCMCs)){
  phybreak.plot.triple(MCMCstate = MCMCs[[i]]$MCMCstate, 
                       title = MCMCs[[i]]$run_name)
}
#if true history is known
for(i in seq_along(MCMCs)){
  phybreak.plot.triple(MCMCstate = MCMCs[[i]]$MCMCstate, 
                       phybreak.true = sim_data[[i]],
                       title = MCMCs[[i]]$run_name)
}
```

## Find effective sample sizes

Since `run.biophybreak` automatically computes and keeps track of effective sample sizes, 
they can be plotted in the following way.

```{r, fig.width=6, fig.height=4}
par(mar = c(5.5, 4, 1, 1))
for(i in seq_along(MCMCs)){
  barplot(log10(MCMCs[[i]]$ESS[[length(MCMCs[[i]]$ESS)]]+1), las = 2)
  abline(h = log10(201))
}
```