---
title: "Biophybreak Workflow"
author: "Erik Lundgren"
date: "November 19, 2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

**biophybreak** can produce probability distributions for infection times based on HIV biomarkers 
as well as infer the transmission history of individuals in a transmission cluster.

### Workflow

- use mbm.predict to find infection age distributions of all individuals
- create a `phybreakdata` object with the sequences and sample times
- create a `phybreak` object with the data, infection and generation time distributions, and inference settings
- run the MCMC without sampling for burn-in
- run the MCMC while sampling for inference
- run desired analysis and plotting

## load packages and set pRNG seed

```{r packages}
library(biophybreak)
library(ggplot2)

seed <- 4321
set.seed(seed)
```

## Using biomarker data to predict infection times

### Create simulated data

We will start by creating some simulated data on which we would like to do inference. 
You would not perform this step if using real datasets.

```{r simulate_data}
#number of individuals
nInd <- 5

#time betweeen infection and sampling is gamma distributed with mean 2 and sd 1.5
infection.ages <- rgamma(nInd, shape = (16/9), rate = (8/9))

#type of distribution for generation function
#(relative chance of transmission based on how long an individual has been infected)
gen.dist.type <- "function"
gen.dist <- stepfun(x = c(0, 0.5), y = c(0, 3, 1)) #three times more likely to transmit in first 6 months

#make a transmission tree
tt <- rtrans_tree(nInd = nInd, 
                  infection.ages = infection.ages, 
                  gen.dist.type = "function", 
                  gen.dist = stepfun(x = c(0, 0.5), y = c(0, 3, 1))) 

#within host model to use
a <- 5 #effective population size
b <- 5 #effective population growth rate per generation (1.5 days)

#mutation model to use
mut.model = "pol" #can be "pol" or "env"

#generate phylogeny and create phybreakdata object
#sim_data <- sim.coal.phybreak(tt = tt, 
#                              a = rep(a, nInd), 
#                              b = rep(b, nInd), 
#                              mut_model = mut.model) 

#load pregenerated sequences for this example
data("example_seqs4321")

#generate phylogeny and create phybreakdata object
sim_data <- sim.coal.phybreak(tt = tt, 
                              a = rep(a, nInd), 
                              b = rep(b, nInd), 
                              user_seqs = example_seqs4321$seqs,
                              user_seq_hosts = example_seqs4321$seq_hosts,
                              user_seq_names = example_seqs4321$seq_names)

#plot simulated transmission history and phylogeny
plotPhyloTrans(sim_data)

#load parameters for multiple biomarker model
data("MBM_pars")

#simluate biomarker values from infection ages
biomarkers <- sim.biomarkers(t.inf = infection.ages, 
                             mub = MBM_pars$mub, #means of MBM parameters
                             Sigmab = MBM_pars$Sigmab, #covariance matrix of MBM parameters
                             sigmae = MBM_pars$sigmae) #standard deviations of gaussian noise for each biomarker
```

### infer infection time distributions with mbm.predict

We can the use the mbm.predict function to infer the infection age distributions for each individual.
You will need JAGS installed for the rjags package to work.

```{r}
infection.dists <- mbm.predict(nInds = 5,
                               BED = biomarkers[,1], #values from BED assay (OD-n))
                               LAg = biomarkers[,2], #value from LAg-Avidity assay
                               CD4 = biomarkers[,3], #CD$+ T cell count
                               pol = biomarkers[,4], #fraction of polymorphic sites in pol gene
                               pol2 = biomarkers[,5], #NGS diversity in pol gene
                               mub = MBM_pars$mub, 
                               Sigmab = MBM_pars$Sigmab, 
                               sigmae = MBM_pars$sigmae,
                               n.adapt = 10000, n.burn = 50000, n.iter = 100000, output.raw = TRUE)
```

## create phybreakdata object

Since we are using simulated data, we already created a `phybreakdata` object with sim.coal.phybreak, 
but with real data, we could create one using the phybreakdata function.

```{r}
data <- phybreakdata(sequences = sim_data$sequences, #sequences
                     sample.times = sim_data$sample.times, #times of samples
                     host.names = sim_data$sample.hosts) #which host each sample was taken from
```

## create phybreak object

```{r}
#set up phybreak object to run MCMC
MCMCstate <- phybreak(dataset = data, 
                      wh.model = "linear", #linear effective population size growth
                      wh.bottleneck = "wide", #wide transmission bottleneck
                      prior.wh.level.mean = 0.1, #pretty conservative; corresponds to a mean of about 24 for a
                      est.sample.mean = FALSE, #do not estimate the sample mean (because we are using the nonparametric distributions)
                      sample.nonpar = TRUE, #we are using the nonparametric distributions from the MBM
                      sample.density = infection.dists$pdf.num[order(tt$t_sam)], #distrubutions from MBM, ordered by sampling time
                      gen.type = gen.dist.type, #kind of generation function
                      gen.density = gen.dist, #generation function
                      post.sam.trans.rate = rep(1, nInd), #no penalty for transmission after sampling
                      est.gen.mean = FALSE, #do not estimate gen mean (because function is provided)
                      use.pml = TRUE, #use the pml function in phangorn for the sequence likelihood with GTR model
                      mut.model = mut.model)
```

## run burn-in

```{r}
MCMCstate <- burnin_phybreak(MCMCstate, ncycles = 10000)
```

## MCMC sampling

```{r}
MCMCstate <- sample_phybreak(MCMCstate, nsample = 25000)
```

## Analysis

### Find effective sample sizes
```{r effective_size}
ESS <- ESS(MCMCstate)
barplot(log10(ESS))
abline(h = log10(200))
```

### MPC tree

Find and plot the maximum parent credibility tree
```{r mpc_tree}
plotPhyloTrans(MCMCstate, plot.which = "mpc")
```

### find posterior probabilities

We can find the posterior probabilities for each infector for each individual (and accuracy if applicable).

```{r}
#since we have simulated data, we can find the accuracy
accuracy <- phybreak.accuracy(phybreak.true = sim_data, MCMCstate = MCMCstate)

#without the simulated data, we can find the posterior pro
post.prob <- phybreak.infector.posts(MCMCstate = MCMCstate)
```

### Plots posterior probabilities

```{r}
#if true history is known
plot(phybreak.plot.posteriors(accuracy$post_prob_df))

#if true history is unknown
plot(phybreak.plot.posteriors(post.prob$post_prob_df))
```